---
title: "Introduction to metaMint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{metaMint-vignette}
  %\VignetteEncoding{UTF-8}{inputenc}
bibliography: references.bib

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",eval=FALSE
)
```


This document introduces to you how to use the package **metaMint** for joint estimation of metabolite and microbial interaction network.

```{r setup}
library(metaMint)
library(dplyr)
library(GGally)
```

## Data

We use an example data set from @mcmillan2015multi that has 51 bacterial taxa and 128 metabolites measured over 131 samples. Taxonomic counts were obtained using 16S rRNA gene sequencing, whereas metabolite profiling was done by gas chromatography-mass spectrometry (GC-MS). 

```{r}
data(BV, package = "metaMint")
```

Before calling the package, it is important to preprocess the taxonomic count data which is subject to the total library size constraint. A common practice is to apply the centered log-ratio (clr) transformation. Briefly, for a vector $r$ with $p$ components, let $g(r)=(\Pi_{j=1}^p)^{1/p}$ denote its geometric mean. The clr of $r$ is defined as
\[
{\rm clr}(r) = \left( \log \frac{r_1}{g(r)},\ldots,\log \frac{r_p}{g(r)}\right)^\intercal.
\]
It is easy to see that the clr transformation does not allow zero counts. To address this issue, a small pseudocount of 0.5 or 1 is often added to all counts before applying the transformation. However, adding pseudocounts may unfairly bias rare species. 

To avoid introducing artificial bias, we use instead the modified-clr (mclr) transformation. The mclr transforms the
non-zero counts with the usual clr and shifts all transformed values to be strictly positive. Details on how mclr compares with clr transformation can be found in @yoon2019microbial. The following Figure illustrates the different marginal distributions one gets with clr and mclr transformation. The mclr transformed value can be modeled using a censored normal distribution, which provides the basis for our method. 

```{r clr,fig.cap='Illustration of how clr and mclr transformation affects the marginal distribution of one OTU.'}
OTUs_clr <- apply(OTUs,2,compositions::clr)
OTUs_mclr <- t(clr.epsilon(t(OTUs)))

ggpairs(data.frame(clr=OTUs_clr[22,],
                   mclr=OTUs_mclr[22,]),
        diag = list(continuous = "barDiag",binwidth = 0.5))
```

## Estimation 

With preprocessed data, we are now ready to show how to use the main functions in **metaMint**. One can use the function `cggm.corr` to estimate the correlations among bacterial taxa.

```{r cor,eval=FALSE}
est_Cor <- cggm.corr(t(OTUs_mclr))
```

Alternatively, if the interest is to learn the conditional association (as defined by partial correlations) among bacterial taxa, one can use the function `cggm.pcorr`. 

```{r invcor,eval=FALSE}
est_InvCor <- cggm.pcorr(t(OTUs_mclr))
```

The model underlying **metaMint** allows for both continuous and zero-inflated variables. Therefore, to jointly estimate the metabolite and microbial interaction network, one only needs to expand the input to include both types of measurements. 

```{r MINT,eval=FALSE}
dat_combined <- t(rbind(OTUs_mclr,metabs))
est_InvCor_combined <- cggm.pcorr(dat_combined)
```

## Model Selection

The function `cggm.stars` provides a way for selecting the optimal regularization parameter that controls the sparsity of the estimated partial correlation network. It is based on the idea of stability selection [@liu2010stability]. 

```{r stars,eval=FALSE}
est_InvCor <- cggm.pcorr(t(OTUs_mclr))
stars_obj <- cggm.stars(est_InvCor,stars.thresh = 0.1, rep.num = 20)
```

## References
